preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-14 Manuscript/manuscript.Rmd
updated figures

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution"
csl: styles.ref/genomebiology.csl
output:
  pdf_document: 
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each set of chromosomes. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. [???update]

@@

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.**(A) The location of preciseTAD (green) vs. Arrowhead (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the predicted probability of each base being a TAD boundary. A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the preciseTAD boundary region (PTBR). (B) Enriched heatmaps comparing the enrichment of CTCF signal around Arrowhead TAD boundaries and preciseTAD boundaries, and (C) around Peakachu loop boundaries and preciseTAD for the entire genome. (D) Log 2 genomic distances from boundaries to nearest CTCF sites. (E) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using Arrowhead; top left), Arrowhead (top right), preciseTAD (using Peakachu; bottom left), and Peakachu (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram.](Figures/figure5.png)

@@

*preciseTAD* offers flexibility in controlling both the number of TAD(loop) boundaries as well as the distance between them. The two primary parameters are the probability threshold (*t*) and $\epsilon$ (refered to as eps; parameter of DBSCAN). The combination of these two quantities change the resulting number of predicted boundaries from *preciseTAD*. A lower threshold and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 4 different thresholds ($t=(0.95,0.975,0.99,1.0)$) and 4 different $\epsilon$ values ($\epsilon=(0.2 r, 0.5r, r, 2r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of t=1.0 and eps=r (Supplementary Figure 9). The same heuristic ($t=1.0$ and $\epsilon=r$) ensured that boundaries were at least 5kb bases away from each other, while also suitably enriched for architectural proteins (Supplementary Figures 10-11). These most optimal heuristics are set as default, but can be further tuned, if necessary.[???update]

@@

**Supplementary Figure 9.** Barplots illustrating the normalized overlap between known molecular drivers of 3D chromatin and flanked predicted TAD boundaries identified by preciseTAD (using Arrowhead ground truth boundaries) for different combinations of thresholds (t) and epsilon parameter values (eps) on GM12878 and for chromosome 22. Normalized overlap was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted (on chr22). Boundaries were flanked by 5 kb resolution.

**Supplementary Figure 10.** Density plots illustrating the genomic distance (in bases) between predicted TAD boundaries identified by preciseTAD (using Arrowhead ground truth boundaries) for different combinations of thresholds (t) and epsilon parameter values (eps) on GM12878 and for chromosome 22. The median genomic distance (range) is provided for each combination.

**Supplementary Figure 11**. The formation of TADs (CHR22: 25mb-27mb) demarcated by boundaries predicted by preciseTAD for different combinations of thresholds (t) and eps values ($\epsilon$) on GM12878. The visualization was created by the juicebox toolbox [@durand2016juicebox] using the GM12878 insitu primary HIC file provided by Rao et al. 2014 [@rao20143d].

>>>>

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution"
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
  pdf_document: 
bibliography: References/references.bib
---

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each set of chromosomes. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B)Bigwig visualization of known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. (C) Enriched heatmaps comparing the enrichment of CTCF signal around ARROWHEAD TAD boundaries and preciseTAD boundaries, and (D) around Peakachu loop boundaries and preciseTAD for the entire genome. (E) Log 2 genomic distances from boundaries to nearest CTCF sites. (F) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using ARROWHEAD; top left), ARROWHEAD (top right), preciseTAD (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram](Figures/figure5.png)

@@

*preciseTAD* offers flexibility in controlling both the number of TAD(loop) boundaries as well as the distance between them. The two primary parameters are the probability threshold (*t*) and $\epsilon$ (refered to as eps; parameter of DBSCAN). The combination of these two quantities change the resulting number of predicted boundaries from *preciseTAD*. A lower threshold and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 4 different thresholds ($t=(0.95,0.975,0.99,1.0)$) and 4 different $eps$ values ($\epsilon=(r, 2r, 3r, 4r, 5r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of t=1.0 and $eps=r$ (Supplementary Figure 9).

@@

**Supplementary Figure 9.** Linecharts illustrating the normalized overlap between known molecular drivers of 3D chromatin and flanked predicted TAD boundaries identified by preciseTAD (using Arrowhead ground truth boundaries) for different combinations of thresholds (t) and epsilon parameter values (eps) on GM12878 and for chromosome 22. Normalized overlap was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted (on chr22). Boundaries were flanked by 5 kb resolution.
